---
title: "Exercises for 02 - Data Import"
format: html
editor: visual
---

## Importing Data from a CSV File

Take a look at the *Files* pane on the bottom right. Within RStudio Cloud, there is a *Cloud* directory that is analogous to a file on your hard drive (but in this case it is on a remote system). The files for this workshop are in a *project* folder, which includes this exercises Quarto file you are working in.

**What folders do you see within the project directory?**

**Click on the *data* directory. How many files are inside it?**

The objective of this exercise is to *import* that CSV file into R as a data frame. Recall that to import CSV files, we use the *read_csv()* function, which is part of the *tidyverse* package. The *read_csv()* function takes a *filename* as an argument and returns a *data frame*. To capture the data frame in a named object, we use the *assignment operator* ( \<- ).

In the following code chunk, we first load the *tidyverse* package. Recall that we load packages the first time we start a new Quarto file, data analysis, or R session. We then use the *read_csv()* function to load the *chem_data.csv* file and capture it in an object named *chem*. Note that the *02 - Data Import.qmd* file is in the *project* directory so we have to provide the *data/* path to point the *read_csv* function to the right location for our file. This step is very important if you follow the best practice of segregating your raw data in a different location from your analysis and output files.

**Run the following code chunk.**

```{r, message = FALSE}
library(tidyverse)
chem <- read_csv("../data/chem_data.csv") # solutions use a different path
```

Hint: click on the green triangle in the upper right corner of the code chunk!

**Stop Here**

## Inspecting a Data Frame

One thing that Excel does well is to provide an interactive visual representation of the data. This allows you to inspect it by sorting and filtering. RStudio actually does this well, too, with one difference - it won't let you *change* any of the data while you inspect it.

Look on the right at the *Environment* pane (**you might have to click on the "Environment" tab**) and find the entry *chem*. This is the data frame you just created inside of R's memory. (If you don't see *chem*, try running the code chunk above again).

**On the far right next to *chem*, you will spot a symbol that looks like a small table. Click on it to make the *Data Viewer* appear.**

Without writing any code, just by working with the Data Viewer, try to answer the following questions:

1.  How many rows are in the data frame? How many columns?

    The data viewer shows the number of rows (observations) and columns (variables) on the bottom - 155544 observations and 13 variables

2.  Go ahead and try to edit one of the values in this viewer. You will find that you can't. It would have been easy for the RStudio programmers to allow editing of specific values, but they decided not to add that feature. Why do you think this was designed that way?

    It is a best practice to not modify your raw data

3.  Each row in *chem* represents a single result from a chemistry test. The value in the *test* column indicates the specific test performed and the value column contains the patient result for that test. What is the largest numerical value in the data set? What is the smallest?

    There is sort functionality that is opened up when you click on the variable names - 0.1 is the smallest value and 16959 is the largest

4.  How many results are available for patients who are 50 and older? Of those patients 50 and older, how many albumin results are available? (Hint: look at the Filter button)

    Clicking the filter button allows you to filter for specific values - 70,320 entries are filtered when you set the age range from 50 to 80

**Stop Here**

## Creating a Quick Cross Tabulation

After reviewing the raw data, it can be helpful to see the relationships between some variables in the form of a contingency table (or cross tabulation). Recall that we installed the *janitor* package in a previous exercise but thus far we have not used any functions from that package. After executing the `install.packages("janitor")` function, we have loaded the package contents onto our (virtual) hard drive. Now let's call the package into our Quarto document and R working session so we can use any functions in the package.

**In the code chunk below use the `library()` function to call the *janitor* package.** Note that once a package is installed onto our system, the quotes around the package name are no longer required. Within the `library()` function the argument will be recognized as an object within R as long as it was already installed.

```{r}
library(janitor)
```

Now that we've loaded the *janitor* package, we can use a function called `tabyl()` to generate a cross tabulation of our *chem* data frame. The `tabyl()` function takes the name of the data frame as its first argument and between 1 and 3 variable (or column) names as subsequent arguments to provide a count of the number of observations (rows) that fall into the categories of the variables you select. This behavior only makes sense with categorical variables. For example, if we want to review the breakdown of observations in our data set between different payor groups (types of insurance), we provide the payor_groups variable as our second argument: (**execute the code by clicking the green play button**)

```{r}
tabyl(chem, payor_groups)
```

Now let's review the counts of observations for both insurance status (variable: *payor_groups*) and patient types (variable: *pat_type*). **Fill in the blanks below to complete this cross tabulation**:

```{r}
tabyl(chem, payor_groups, pat_type)
```

This function can be helpful if you have a specific question regarding the frequency of categorical observations and can overcome the limitations of the `summary()` function if data is not the expected type (e.g. categorical variables are not factors). We will consider data types in more detail in the next lesson.
